Appendix: Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-02-27 Appendix_plain_v2.tex
push ole and owen edits to DAG and appendix.

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA $copies$ $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Let $\alpha_0$ be the intercept representing the mean hake DNA concentration and $\epsilon(x,y,d)$ be a random effect that allows for the effect of spatial location and water depth. We write these as additive on the log-scale,

\begin{align}
  D_H(x,y,d) &= \psi_0 + \epsilon(x,y,d) \\
  \epsilon(x,y,d) &\sim N(0,\sigma_D)
\end{align}

This model describes perhaps the simplest possible random effect model and notably does not leverage any information about the spatial arrangement of samples (i.e. the random effect does not include a spatial correlation component).  This is intentional.  The focus of this model is on the other species DNA concentration (not Pacific hake) and therefore we allow the estimated hake concentrations to follow the qPCR observations as much as possible.

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu L$ aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see Shelton et al. 2022 Supp S1 for additional description).

Finally, we add a random effect for the individual bottles sampled at each year-station-depth combination ($\delta_i(x,y,d)$). This effect  describes the deviation of individual sampled bottles from the location mean, $\delta_i(x,y,d) \sim N(0,\tau)$, with $\tau$ indicating a depth-specific standard deviation among bottles. Because we have only two replicate bottles at each location, we impose a sum to zero constraint on the bottles collected from a single location (i.e. for a given location, $\sum_i \delta_i(x,y,d) = 0$). Then, the log-concentration of hake DNA in a sample analyzed by qPCR, $E_{Hi}$ is,

\begin{align}
  E_{Hi} &= D_H({x,y,d}) +  \log{V_i} + \log{I_i} + \mathbf{I}\zeta + \delta_i(x,y,d)
\end{align}

We connect the estimated concentrations to observations from qPCR using two likelihoods connected to the concentration $E$. First we determine whether amplification was detected in each qPCR replicate $r$ run on each plate $j$, $Z_{Hijr}$ and if amplification was observed we model the PCR cycle at which amplification was detected $Y_{Hijr}$; $Y_{Hijr}$ is observed as a continuous, positive value. DNA concentrations have a log-linear relationships with $Y_{Hijr}$; smaller values of $Y_{Hijr}$ are associated with higher DNA concentrations.

Because we are modeling the discrete number of DNA molecules that are present in the assayed sample, we know that the Poisson distribution provides an appropriate observation distribution for the number of molecules present in a given qPCR reaction. Assuming a Poisson distribution, conditional on a true mean number of DNA copies in the sample $X$, the probability of having exactly zero DNA copies in a qPCR reaction is $e^{-X}$ and the probability of having non-zero DNA copies is the complement, $1-e^{-X}$. If there are exactly zero molecules in a qPCR reaction, amplification and therefore detection of amplification will not occur. However, there are other factors (e.g. PCR inhibition) that may reduce amplification further and therefore we expect the probability of amplification to be at most $1-e^{-X}$. As a result, we estimate an additional term, $\phi$ representing the fractional reduction in amplification efficiency due to other factors on qPCR plate $j$ ($0<\phi<1$). Then we have a pair of observation models for hake DNA, both of which are a function of DNA concentration, $E_{Hi}$

\begin{align}
  Z_{Hijr} &\sim Bernoulli\left(1 - \exp(-2 e^{E_{Hij}}\phi) \right) \\ 
  Y_{Hijr} &\sim Normal \left(\beta_{0j}+\beta_{1j}{E_{Hij}},\sigma_Y(E_{Hij}) \right) \qquad  if \: Z_{Hijr} = 1 
\end{align}

The 2 is present in the first line of the equation because we use 2 $\mu L$ of sample in each qPCR reaction and so the expected DNA copies in reaction is $2e^{E_{ij}}$. We allow $\sigma_C(E_{ij})$ to vary as a log-linear function of DNA concentration to account for the fact that there is decreased variability in qPCR measures of $C_t$ at higher DNA concentrations: $\sigma_C({E_{ij}}) = exp(\gamma_{0}+\gamma_{1}E_{ij})$ (see also Fig. XX).

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu L^{-1}$, each of 2$\mu L$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{jp}$ be the known copy number for sample $j$ in PCR plate $p$, then,

\begin{align}
  Z_{jp} &\sim Bernoulli \left(1-exp(-2 K_{jp} \phi) \right)\\
  Y_{jp} &\sim Normal \left(\beta_{0j}+\beta_{1j}\log{K_{jp}},\sigma_Y(K_{jp})\right) \quad  if \: Z_{jp} =1
\end{align}

Note that there are different intercept $\beta_{0p}$ and slope $\beta_{1p}$ parameters for each PCR plate to allow for among-plate variation in amplification. We model each calibration paratemeter ($\beta_{0p},\beta_{1p})$ hierarchically using a normal distribution, with among plate mean and variance (i.e. $\beta_{0p} \sim N(\mu_{\beta_{0}},\sigma_{\beta_{0}})$).

@@

In addition to the qPCR observations, we used the MiFish primer and high throughput sequencing to detect and enumerate the sequences of other fish species in the same samples that were examined for hake DNA. Again, we are interested in determining the DNA concentration of each focal species in each sample. Unfortunately, metabarcoding data is compositional data (see *Glor et al. *, *Shelton et al. 2023*), and so while it can inform the relative abundance of species within a sample, alone it cannot provide estimates of DNA concentration. Fortunately, the qPCR data from Pacific hake provides information about DNA concentration for one species and the metabarcoding data provides information about the relative abundance for tens of species, including hake. Here, we follow and extend the metabarcoding model developed by Shelton et al. (2023) and show how it links with the qPCR model above.

As above, we let $D_s(x,y,d)$ be the log-scale eDNA concentration but now for species $s$. We let $E_{si}(x,y,d)$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; *Aitchison XXXX*).  We define a reference species in each sample, $R$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see Shelton et al. 2023).  We can then calculate $\nu_{si}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see *Mock Communities* below),

\begin{align} 
  \nu_{si} &= \left( E_{si}(x,y,d) - E_{Ri}(x,y,d)\right) + N_{PCR} \left(\alpha_s - \alpha_R \right) 
\end{align}

To maintain identifiability, we assign one $\alpha_s = 0$. We can then convert the $\nu$s to the predicted proportions of the metabarcoding reads $\mu_{si}$ using the softmax transformation and to the observed reads via a Dirichlet-Multinomial likelihood,

\begin{align} 
  \pi_{si} &= \dfrac{e^{\nu_{si}}}{\sum_{s=1}^S e^{\nu_{si}}} \\
  \boldsymbol{R}_{i} & \sim Multinomial\left(\boldsymbol{\pi}_{i},T_{i}\right) 
\end{align}

where $T_{i}$ is the total number of reads observed across all species for that bottle.

As with the qPCR analyses, in the absence of additional information estimates of $\alpha$ and $E$ cannot be statistically identified.  We use the mock community data, which has known input DNA concentrations and  same basic model as for the metabarcoding observations but for the moe

@@

\textbf{OWEN.  COULD YOU HELP HERE?}

The above statistical model provides predictions of species-specific DNA concentrations (D) at each depth sampled (d = 0, 50, 150, 300 or 500 m). We used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA , as:
$$\ln D_{xyz} = \tau + \eta(s) + \varepsilon_{d}(s)$$
where x,y,z index longitude, latitude, and depth, respectively, $\tau$ is scalar intercept, $\eta(s)$ is a spatial smooth shared among depths, and $\varepsilon_d(s)$ is a depth-specific spatial smooth. Both $\eta(s)$ and $\varepsilon_d(s)$ are a zero-centered Gaussian random field estimated using the stochastic partial differntial equation (SPDE) approximation (see \cite{anderson2022}). Each species was fitted individually while we created and selected the best mesh for approximating spatial Gaussian random fields as part of the model.

\subsection*{Model convergance}
OLE + GLED

>>>>

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA copies $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Hereafter we refer to a combination of spatial location and depth as a ``station" and index station using $j$ (thus, $D_{Hj} = D_H(x,y,d)$). Let $\psi$ be the intercept representing the mean hake DNA concentration and $\epsilon_j$ be a random effect that allows for the effect of that station. We write these as additive on the log-scale,

\begin{align}
  D_{Hj} &= \psi + \epsilon_j \\
  \epsilon_j &\sim  \mathrm{Normal}(0,\tau)
\end{align}
This model describes perhaps the simplest possible random effect model and notably does not leverage any information about the spatial arrangement of samples (i.e. the random effect does not include a spatial correlation component).  This is intentional.  The focus of this model is on the other species DNA concentration (not Pacific hake) and therefore we allow the estimated hake concentrations at a to follow the qPCR observations as much as possible.

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu L$ aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see Shelton et al. 2022 Supp S1 for additional description). Here we fix $\zeta$ = \textbf{XXX} based on previous analyses (Shelton et al. 2022).

Finally, we add a random effect for the individual bottles sampled at each station ($\delta_{ij}$). This effect  describes the deviation of individual sampled bottles from the location mean, $\delta_{ji} \sim  \mathrm{Normal}(0,\sigma_D)$, with $\sigma_D$ indicating a  standard deviation among bottles. Because we have only two replicate bottles at each location, we impose a sum to zero constraint on the bottles collected from a single location (i.e. for a given location, $\sum_i \delta_{ji} = 0$). Then, the log-concentration of hake DNA in a sample analyzed by qPCR, $E_{Hji}$ is,

\begin{align}
  E_{Hji} &= D_{Hj} +  \log{V_i} + \log{I_i} + \mathbf{I}\zeta + \delta_{ji}
\end{align}
We connect the estimated concentrations to observations from qPCR using two likelihoods connected to the concentration $E$. First we determine whether amplification was detected in each qPCR replicate $r$ run on each plate $p$, $Z_{Hjipr}$ and if amplification was observed we model the PCR cycle at which amplification was detected $Y_{Hjipr}$; $Y_{Hjipr}$ is observed as a continuous, positive value. DNA concentrations have a log-linear relationships with $Y_{Hjipr}$; smaller values of $Y_{Hjipr}$ are associated with higher DNA concentrations.

Because we are modeling the discrete number of DNA molecules that are present in the assayed sample, we know that the Poisson distribution provides an appropriate observation distribution for the number of molecules present in a given qPCR reaction. Assuming a Poisson distribution, conditional on a true mean number of DNA copies in the sample $X$, the probability of having exactly zero DNA copies in a qPCR reaction is $e^{-X}$ and the probability of having non-zero DNA copies is the complement, $1-e^{-X}$. If there are exactly zero molecules in a qPCR reaction, amplification and therefore detection of amplification will not occur. However, there are other factors (e.g. PCR inhibition) that may reduce amplification further and therefore we expect the probability of amplification to be at most $1-e^{-X}$. As a result, we estimate an additional term, $\phi$, representing the fractional reduction in amplification efficiency due to other factors during qPCR ($0<\phi<1$). Then we have a pair of observation models for hake DNA, both of which are a function of DNA concentration, $E_{Hji}$,

\begin{align}
  Z_{Hjipr} &\sim  \mathrm{Bernoulli} \left(1 - \exp(-2 e^{E_{Hji}}\phi) \right) \\ 
  Y_{Hjipr} &\sim  \mathrm{Normal} \left(\beta_{0p}+\beta_{1p}{E_{Hji}},\sigma_{Yji} \right) \qquad   \mathrm{if} \: Z_{Hjipr} = 1 
\end{align} 
The 2 is present in the first line of the equation because we use 2 $\mu$L of sample in each qPCR reaction and so the expected DNA copies in reaction is $2e^{E_{Hji}}$. We allow $\sigma_{Yji}$ to vary as a log-linear function of DNA concentration to account for the fact that there is decreased variability in qPCR measures of $C_t$ at higher DNA concentrations: $\sigma_{Yji} = exp(\gamma_{0}+\gamma_{1}E_{Hji})$.

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu L^{-1}$, each of 2$\mu L$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{qp}$ be the known log-copy number for sample $q$ in PCR plate $p$, then,

\begin{align}
  Z_{qpr} &\sim  \mathrm{Bernoulli} \left(1-exp(-2 e^{K_{qp}} \phi) \right)\\
  Y_{qpr} &\sim  \mathrm{Normal} \left(\beta_{0j}+\beta_{1j}{K_{qp}},\sigma_{Yqp}\right) \quad  \mathrm{if} \: Z_{qjp} =1
\end{align}
Note that there are different intercept ($\beta_{0p}$) and slope ($\beta_{1p}$) parameters for each PCR plate to allow for among-plate variation in amplification. We model each calibration parameter ($\beta_{0p},\beta_{1p})$ hierarchically using a normal distribution, with among plate mean and variance (i.e. $\beta_{0p} \sim \mathrm{Normal}(\mu_{\beta_{0}},\sigma_{\beta_{0}})$).

@@

In addition to the qPCR observations, we used the MiFish primer and high throughput sequencing to detect and enumerate the sequences of other fish species in the same samples that were examined for hake DNA. Again, we are interested in determining the DNA concentration of each focal species in each sample. Unfortunately, metabarcoding data is compositional data \cite{shelton2023}, and so while it can inform the relative abundance of species within a sample, alone it cannot provide estimates of DNA concentration. Fortunately, the qPCR data from Pacific hake provides information about DNA concentration for one species and the metabarcoding data provides information about the relative abundance for tens of species, including hake. Here, we follow and extend the metabarcoding model developed by Shelton et al. (2023) and show how it links with the qPCR model above.

As above, we let $D_{sj}$ be the log-scale eDNA concentration but now for species $s$. We let $E_{sji}$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; \textbf{Aitchison 1986}).  We define a reference species in each sample, $R_i$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see \cite{shelton2023}).  We can then calculate $\nu_{sji}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see \textit{Mock Communities} below),

\begin{align} 
  \nu_{sji} &= \left( E_{sji} - E_{Rji}\right) + N_{PCR} \left(\alpha_s - \alpha_{R_i} \right) 
\end{align}
We can then convert the $\nu$s to the predicted proportions of the metabarcoding reads $\pi_{sji}$ using the softmax transformation and to the observed reads via a multinomial likelihood,

\begin{align} 
  \pi_{sji} &= \dfrac{e^{\nu_{sji}}}{\sum_{s=1}^S e^{\nu_{sji}}} \\
  \boldsymbol{W}_{ji} & \sim  \mathrm{Multinomial}\left(\boldsymbol{\pi}_{ji},T_{i}\right) 
\end{align}
where $T_{i}$ is the total number of reads observed across all species for that bottle.

As with the qPCR analyses, in the absence of additional information estimates of $\alpha$ and $E$ cannot be statistically identified.  We use the mock community data, which has known input DNA concentrations and  same basic model as for the metabarcoding observations but instead of having unknown DNA concentrations, we have known inputs for each species $L_{sm}$ in mock community $m$.

\begin{align} 
 \nu_{sm} &= \left( L_{sm} - L_{Rm}\right) + N_{PCR} \left(\alpha_s - \alpha_{R_i} \right) \\
  \pi_{sm} &= \dfrac{e^{\nu_{sm}}}{\sum_{s=1}^S e^{\nu_{sm}}} \\
  \boldsymbol{W}_{m} & \sim  \mathrm{Multinomial}\left(\boldsymbol{\pi}_{m},T_{m}\right) 
\end{align}
To maintain identifiability, we assign one $\alpha_s = 0$.

@@

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). To generate smooth distribution models, we use posterior means of $E_s(x,y,d)$ from the joint statistical model, exponentiated it to move to concentration from log concentration, $G_{s}(x,y,d) = exp(E_{s}(x,y,d))$, and used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA. Because of the predominance of very low estimated DNA concentrations at a large number of site, we set $G_{s}(x,y,d) < 0.5$ equal to zero and then used a Tweedie likelihood in sdmTMB,

\begin{align} 
	 log(F_s(x,y,d)) = \kappa + \eta(x,y) + \varepsilon_{d}(x,y,d) \\
	 G_{s}(x,y,d) \sim \mathrm{Tweedie}(F_s(x,y,d),p ,\theta_s)
\end{align}
where x, y, and d index longitude, latitude, and depth, respectively.   $\kappa$ is scalar intercept. Both $\eta(x,y)$ and $\varepsilon_d(x,y,d)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim  \mathrm{MVN}(0,\sum_\eta)$ and $\varepsilon_d \sim  \mathrm{MVN}(0,\sum_{\varepsilon_d})$ (see \cite{anderson2022}).  $\eta$ is a depth-invariant spatial field, while $\varepsilon_d$ is a spatial field specific to each depth category. In simple terms, each station-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated DNA, while $\varepsilon_d$ modulates that estimate for different sampling depths. The $\varepsilon_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$.

Estimation of large, multi-level spatiotemporal models can strain computational power, in part because of the need in spatial models to invert large covariance matrices between the data. We use the Stochastic Partial Differential Equation (SPDE) model, implemented through the fmesher R package (Lindgren et al. 2024; 10.32614/CRAN.package.fmesher) to approximate the latent spatial process and its variance by using a triangulated mesh (Lindgren et al. 2011). The SPDE approach drastically reduces computation time by using a sparse covariance matrix to approximate Gaussian Markov random fields like $\eta$ and $\varepsilon_d$. Model estimates and the degree of spatial smoothing can be affected by the details of mesh construction, and we therefore produced a range of meshes with different complexities (i.e., varying numbers of triangles, triangle sizes, and therefore vertices) to test for each species.

Each species was fitted individually using the equation above. We tested 18 candidate meshes for each species, increasing in complexity from 33 to 205 vertices, as well as allowing each model to include or exclude the scalar intercept $\kappa$. Each candidate model for each species was fit by maximum marginal likelihood, implemented in the sdmTMB package in R ( \cite{anderson2022} ). Model selection was carried out by first ensuring proper convergence, then choosing the model for each species that maximized the summed likelihood.

We present results based on the predicted smooth mean concentration, $F_s(x,y,d)$, in all results.

\subsection*{Model convergance}
